\documentclass[11pt]{amsart}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1.2in]{geometry}
\usepackage{booktabs}
\usepackage{adjustbox}
\usepackage{xcolor}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

% Last-resort stretch so a paragraph that cannot break cleanly sets loose
% rather than running into the right margin.
\setlength{\emergencystretch}{2em}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{question}[theorem]{Question}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\FF}{\mathbb{F}}
\newcommand{\NN}{N}
\newcommand{\MM}{M}
\newcommand{\conv}{\operatorname{conv}}
\newcommand{\Cone}{\operatorname{Cone}}
\newcommand{\Sing}{\operatorname{Sing}}
\newcommand{\mult}{\operatorname{mult}}
\newcommand{\vertt}{\operatorname{vert}}

\title[Non-smoothability and mirror non-isolatedness]{Non-smoothability and
mirror non-isolatedness\\ of Batyrev Calabi--Yau threefolds}
\author{Bernd J. Wuebben}
\date{July 14, 2026}

\begin{document}

\begin{abstract}
Let $\Delta \subset \ZZ^4$ be a reflexive polytope with unit edges, so that the
generic anticanonical hypersurface $X \subset \PP_\Delta$ is a Calabi--Yau
threefold whose singular points are the cones over the two-dimensional faces of
$\Delta$.  We ask when the Batyrev mirror $X^\circ \subset \PP_{\Delta^\circ}$ is
\emph{also} isolated-singular, and relate this to the local models of the
prequels.  We first record an exact identity: the number $\ell(F^\circ)$ of
singular points of $X$ over a two-dimensional face $F$ equals the multiplicity of
the transverse cone of $F$; consequently both $X$ and $X^\circ$ are
isolated-singular --- we call $\Delta$ \emph{both-sides unit} --- if and only if
every two-dimensional face of $\Delta$ is transversally smooth.  We then argue
that the novel non-smoothable germs of the prequels --- the anticanonical cone
over $\FF_1$, and the deformable-but-non-smoothable pentagon cone --- cannot occur
on a both-sides unit polytope.  Precisely, we conjecture that \emph{every
two-dimensional face of a both-sides unit reflexive $4$-polytope is a triangle or
a centrally-symmetric polygon}, and we support this by a complete check of the
$3{,}905{,}789$ reflexive $4$-polytopes with at most nine vertices and of the
non-smoothable-face consequence for the $3{,}735{,}806$ polytopes with at
least $23$ vertices, by the
explicit structure of the both-sides examples --- including the $36$-vertex
hexagon-product polytope at the top of the classification, which is
both-sides unit, conforms, and mirrors a smooth Calabi--Yau threefold ---
and by a reduction of the
conjecture to a purely local lemma about a face and its star.  The conjecture
implies that the Batyrev mirror of a Calabi--Yau threefold carrying a
non-quotient or a deformable non-smoothable point is always non-isolated, so that
its singularities leave the isolated Altmann--Corti--Filip--Petracci framework
and are governed instead by Filip's non-isolated $0$-mutability criterion.  All
computations are performed by the exact-arithmetic programs of the prequels,
extended by a scanner included as ancillary files.
\end{abstract}

\maketitle

\section{Introduction}\label{sec:intro}

Let $\Delta \subset \NN_\RR = \RR^4$ be a reflexive polytope all of whose edges
have lattice length one, let $\Delta^\circ \subset \MM_\RR$ be its polar dual,
and let $X \in |{-K_{\PP_\Delta}}|$ be a generic anticanonical hypersurface in the
Gorenstein Fano toric fourfold associated to the face fan of $\Delta$.  By
Batyrev \cite{Batyrev94} $X$ is a Calabi--Yau threefold with Gorenstein canonical
singularities, and by the planting proposition of \cite{paper1} its singular
locus is exactly: for each two-dimensional face $F \preceq \Delta$ whose cone
$C(F)$ is singular, $\ell(F^\circ)$ points at which the germ of $X$ is
analytically isomorphic to $C(F)$, where $\ell(F^\circ)$ is the lattice length of
the edge $F^\circ$ of $\Delta^\circ$ dual to $F$.  The trichotomy of \cite{paper1}
classifies $C(F)$ by the combinatorics of $F$ into rigid \textup{(R)},
deformable-but-non-smoothable \textup{(D)}, and smoothable \textup{(S)}; the first
two are non-smoothable, and a single such face forces $X$ to admit no smoothing.

Mirror symmetry, in the Batyrev picture, is the polar duality
$\Delta \leftrightarrow \Delta^\circ$, which for $4$-polytopes exchanges
two-dimensional faces with edges.  The mirror family is the generic anticanonical
$X^\circ \subset \PP_{\Delta^\circ}$, whose singular locus is governed in the
same way by the two-dimensional faces of $\Delta^\circ$.  The purpose of this note
is to understand what the non-smoothability of $X$ --- that is, the presence of a
face of type \textup{(R)} or \textup{(D)} --- means for $X^\circ$, and in
particular when $X$ and $X^\circ$ are \emph{both} isolated-singular.

Our first observation is exact and elementary.  For a two-dimensional face $F$
write $u_1, u_2 \in \MM$ for the primitive inner normals of the two facets of
$\Delta$ containing $F$, so $F^\circ = \conv\{u_1,u_2\}$, and let
$\MM_F = \MM \cap (\RR u_1 + \RR u_2)$, a rank-two saturated sublattice.

\begin{theorem}[= Lemma~\ref{lem:mult}]\label{thm:intromult}
$\ell(F^\circ) = [\,\MM_F : \ZZ u_1 + \ZZ u_2\,]$, the multiplicity of the
transverse cone $\Cone(u_1,u_2)$ of $F$.  In particular $\ell(F^\circ)=1$ if and
only if $u_1, u_2$ form a basis of $\MM_F$, i.e.\ the transverse cone is smooth.
\end{theorem}

Since $X$ is isolated-singular exactly when $\Delta$ has unit edges, and
$X^\circ$ exactly when $\Delta^\circ$ does --- equivalently when every
two-dimensional face $F$ of $\Delta$ has $\ell(F^\circ)=1$ --- Theorem
\ref{thm:intromult} gives a clean reading of the doubly-isolated condition.

\begin{corollary}\label{cor:introboth}
$X$ and $X^\circ$ are both isolated-singular if and only if $\Delta$ has unit
edges and every two-dimensional face of $\Delta$ is transversally smooth.  We
call such $\Delta$ \emph{both-sides unit}.
\end{corollary}

The point-packets of $X$ and the singular curves of $X^\circ$ correspond exactly:
the $\ell(F^\circ)$ points of type $C(F)$ over $F$ are mirror to a single curve of
transverse $A_{\ell(F^\circ)-1}$ singularities on $X^\circ$ along the toric curve
dual to $F$ (\S\ref{sec:corr}).  Thus $X^\circ$ is isolated precisely when no such
curve appears, i.e.\ on a both-sides unit polytope.

Our main assertion is that the non-smoothable local models introduced in the
prequels are incompatible with this doubly-isolated situation, except in the
classical quotient case.  Call a lattice polygon a \emph{zonotope} if it is a
Minkowski sum of lattice segments; a unit-edge polygon is a zonotope if and only
if it is centrally symmetric, and every such polygon is smoothable of type
\textup{(S)}.

\begin{conjecture}[triangle-or-zonotope]\label{conj:main}
Every two-dimensional face of a both-sides unit reflexive $4$-polytope is either a
triangle or a zonotope.
\end{conjecture}

Equivalently: on a both-sides unit polytope every non-smoothable two-dimensional
face is a triangle, hence a cyclic quotient singularity.  We are unable to prove
Conjecture~\ref{conj:main}, but we give four kinds of evidence
(\S\ref{sec:conj}): a complete verification over the
$3{,}905{,}789$ reflexive $4$-polytopes with at most nine vertices (exactly
five both-sides unit polytopes, all conforming); a verification of its
non-smoothable-face consequence over the $3{,}735{,}806$ polytopes with at
least $23$ vertices, where the sweep of the prequels recorded no both-sides
polytope carrying a non-smoothable face at all; the explicit
mirror-pair structure of the both-sides examples --- including a new pair far
beyond the low-vertex range, the $36$-vertex product of two reflexive
hexagons (the largest reflexive $4$-polytope by vertex count) and its
$12$-vertex polar, whose faces are zonotopes and smooth triangles
respectively; and a reduction of the
conjecture to a purely local statement about a face and its star, verified on
its $38{,}571$ candidate instances over the polytopes with at most eight
vertices.  We also explain why the natural stronger guess --- that
a both-sides unit polytope is simplicial --- is \emph{false}, by a self-duality
argument (Remark~\ref{rem:selfdual}).

The geometric consequence is the following dichotomy, which we regard as the main
message of the note; it is unconditional on the quotient side and conditional on
Conjecture~\ref{conj:main} on the other.

\begin{corollary}[mirror non-isolatedness]\label{cor:mirror}
Let $\Delta$ be reflexive with unit edges and let $X$ have a non-smoothable
singular point whose germ is \emph{not} a cyclic quotient singularity --- for
instance the anticanonical cone over $\FF_1$ of \cite[Thm.~5.1]{paper1}, or the
deformable-but-non-smoothable pentagon cone of \cite[Thm.~5.3]{paper1}.  Then,
assuming Conjecture~\ref{conj:main}, the Batyrev mirror $X^\circ$ is
non-isolated.  Consequently the mirror singularities of the novel non-smoothable
Calabi--Yau threefolds of \cite{paper1} always lie outside the isolated
Altmann--Corti--Filip--Petracci framework and are governed by Filip's
non-isolated $0$-mutability criterion \cite{Filip25}.
\end{corollary}

A clean two-sided, isolated-to-isolated correspondence between non-smoothable
Calabi--Yau threefolds therefore exists \emph{only} in the classical quotient
(triangle) stratum, and even there the both-sides examples pair a singular
threefold with a smooth one rather than singular with singular
(Example~\ref{ex:pair}).  The natural home of the mirror question for the new
phenomena is thus the non-isolated theory of \cite{Filip25}, and, conjecturally,
the mirror-theoretic language of $0$-mutable Laurent polynomials
\cite{CFP} in which non-smoothability should read as a combinatorial obstruction
(\S\ref{sec:questions}).

All computations in this note --- the transverse multiplicities, the both-sides
scans of the Kreuzer--Skarke list (a reference scanner and a fast-engine
scanner whose rare hits are re-verified by the reference toolkit), the
$36$-vertex mirror pair, and the local reduction --- are implemented in
the exact-arithmetic programs of \cite{paper1}, extended by the scanners
described in \S\ref{sec:conj}; they are included with the arXiv submission as
ancillary files.

\subsection*{Acknowledgements}
This note continues \cite{paper1,paper2}; the question of smoothing canonical
Calabi--Yau singularities was posed to the author by Mark Gross many years ago.

\section{Preliminaries}\label{sec:prelim}

We keep the notation of \cite{paper1}.  For a lattice polygon $Q$ with unit edges
placed at height one, $C(Q)$ denotes the associated Gorenstein toric affine
threefold; $Q$ is of type \textup{(R)}, \textup{(D)} or \textup{(S)} according as
its edge-vector multiset has no proper zero-sum sub-multiset, has one but no
partition into unit segments and unimodular triples, or has such a partition
\cite{Altmann97,CFP}.  For a reflexive polytope $\Delta$ we write $\Delta^\circ$
for the polar dual; a face $G \preceq \Delta$ has a dual face
$G^\circ \preceq \Delta^\circ$ with $\dim G + \dim G^\circ = 3$, so
two-dimensional faces $F$ of $\Delta$ are dual to edges $F^\circ$ of
$\Delta^\circ$ and vice versa.  We write $\ell(\cdot)$ for lattice length.  By the
planting proposition \cite[Prop.~3.1]{paper1}, for $\Delta$ with unit edges the
generic $X \in |{-K}|$ meets the toric curve $V(\sigma_F)$ of a two-dimensional
face $F$ in $\ell(F^\circ)$ reduced points, at each of which the germ of $X$ is
$C(F)$; $X$ is smooth away from these points, so $X$ is isolated-singular exactly
when $\Delta$ has unit edges.

\section{The transverse-multiplicity identity}\label{sec:mult}

Let $F$ be a two-dimensional face of a reflexive polytope $\Delta \subset
\NN_\RR$, and let $u_1, u_2 \in \MM$ be the primitive inner normals of the two
facets of $\Delta$ containing $F$, so that $\langle u_i, \cdot\rangle = -1$ on
$F$ and $F^\circ = \conv\{u_1, u_2\}$.  Put $\MM_F = \MM \cap (\RR u_1 + \RR u_2)$,
a rank-two saturated sublattice of $\MM$ containing $u_1, u_2$ as primitive
vectors.

\begin{lemma}\label{lem:mult}
$\ell(F^\circ) = [\,\MM_F : \ZZ u_1 + \ZZ u_2\,]$.
\end{lemma}

\begin{proof}
Both sides depend only on the pair $(u_1,u_2)$ inside the rank-two lattice
$\MM_F$.  Choose a $\ZZ$-basis of $\MM_F$ in which $u_1 = (0,-1)$; this is
possible because $u_1$ is primitive.  Write $u_2 = (p,q)$; then $p =
\det(u_1,u_2) = \pm[\,\MM_F : \ZZ u_1 + \ZZ u_2\,]$, and $\ell(F^\circ) =
\ell\bigl(\conv\{u_1,u_2\}\bigr) = \gcd(u_2 - u_1) = \gcd(p,\,q+1)$.

Let $v$ be a vertex of $F$ and let $\bar v \in \NN_F := \NN/(\NN \cap u_1^\perp
\cap u_2^\perp)$ be its image; since $u_1,u_2$ vanish on $\NN \cap u_1^\perp \cap
u_2^\perp$ they descend to the dual basis of $\MM_F = \NN_F^\vee$, and $\bar v$ is
a lattice point of $\NN_F \cong \ZZ^2$ with $\langle u_i, \bar v\rangle = -1$.
Writing $\bar v = (a,b)$ in the dual basis, $\langle u_1, \bar v\rangle = -b = -1$
gives $b = 1$, and $\langle u_2, \bar v\rangle = pa + q = -1$ gives $p a = -1-q$,
so $p \mid (q+1)$.  Hence $\gcd(p, q+1) = |p|$, and
$\ell(F^\circ) = |p| = [\,\MM_F : \ZZ u_1 + \ZZ u_2\,]$.
\end{proof}

\begin{remark}
Lemma~\ref{lem:mult} is elementary and may be known; what matters here is the
reading it provides.  The right-hand side is the multiplicity of the
two-dimensional cone $\Cone(u_1,u_2)$, which is the normal cone of $F$ and dually
the transverse cone of $\Delta$ along $F$ (the image of the tangent cone of
$\Delta$ at an interior point of $F$ under the quotient $\NN_\RR \to \NN_\RR /
\operatorname{lin}(F)$).  Thus \emph{the dual-edge length counts exactly the
transverse singularity multiplicity}, and $\ell(F^\circ)=1$ is the statement that
$\Delta$ is transversally smooth along $F$.  The lattice-point hypothesis on $v$ is
essential: for an arbitrary pair of primitive vectors $u_1,u_2$ the two sides of
Lemma~\ref{lem:mult} differ.
\end{remark}

Corollary~\ref{cor:introboth} is immediate: $X$ is isolated-singular iff $\Delta$
has unit edges (\S\ref{sec:prelim}); $X^\circ$ is isolated-singular iff
$\Delta^\circ$ has unit edges, i.e.\ every edge $F^\circ$ of $\Delta^\circ$ has
$\ell(F^\circ)=1$, i.e.\ by Lemma~\ref{lem:mult} every two-dimensional face of
$\Delta$ is transversally smooth.

\section{The mirror correspondence of the singular strata}\label{sec:corr}

We record the mirror-dual description of the singular loci, which puts
Corollary~\ref{cor:introboth} in geometric terms.  It is a direct consequence of
Batyrev's construction and the planting proposition, and was the starting point of
this investigation.

\begin{proposition}\label{prop:corr}
Let $\Delta$ be reflexive with unit edges and $F$ a two-dimensional face with
$C(F)$ singular.  Under polar duality $F$ corresponds to the edge $F^\circ$ of
$\Delta^\circ$, and:
\begin{enumerate}
\item $X$ has $\ell(F^\circ)$ isolated points of type $C(F)$ over $F$;
\item $X^\circ$ has, along the toric curve $V(\sigma_{F^\circ})$ dual to
$F^\circ$, a curve of transverse $A_{\ell(F^\circ)-1}$ singularities (and none if
$\ell(F^\circ)=1$).
\end{enumerate}
In particular $X^\circ$ is isolated-singular if and only if
$\max_F \ell(F^\circ) = 1$, i.e.\ $\Delta$ is both-sides unit.
\end{proposition}

\begin{proof}
(1) is \cite[Prop.~3.1]{paper1}.  For (2), the edge $F^\circ$ of $\Delta^\circ$
has lattice length $\ell(F^\circ)$, so it carries $\ell(F^\circ)-1$ interior
lattice points; the corresponding toric curve $V(\sigma_{F^\circ}) \subset
\PP_{\Delta^\circ}$ meets the generic anticanonical $X^\circ$ in a
$1$-dimensional stratum along which $\PP_{\Delta^\circ}$ has a transverse
$A_{\ell(F^\circ)-1}$ singularity (an edge of lattice length $m$ produces a curve
of transverse $A_{m-1}$ singularities, \cite[\S2]{paper1}).  Hence $X^\circ$ is
smooth in codimension two along $V(\sigma_{F^\circ})$ precisely when
$\ell(F^\circ)=1$, and it is isolated-singular iff this holds for every
two-dimensional face $F$, i.e.\ iff $\Delta$ is both-sides unit.
\end{proof}

Thus the point-packet of $X$ over $F$ and the $A_{\ell(F^\circ)-1}$-curve of
$X^\circ$ are mirror-dual, and the two threefolds are simultaneously
isolated-singular exactly on the both-sides unit polytopes.

\section{Both-sides unit polytopes and the triangle-or-zonotope
conjecture}\label{sec:conj}

We now study which local models can occur on a both-sides unit polytope.  Recall
that a unit-edge polygon is a \emph{zonotope} --- a Minkowski sum of unit
segments --- if and only if it is centrally symmetric (its edge vectors occur in
antipodal pairs), and that every zonotope is of type \textup{(S)}: the antipodal
pairs partition the edge multiset into unit segments, giving a smoothing.  Thus
Conjecture~\ref{conj:main} asserts that the only non-smoothable faces of a
both-sides unit polytope are triangles.

\begin{remark}[why not ``simplicial'']\label{rem:selfdual}
It is tempting to conjecture the stronger statement that every two-dimensional
face of a both-sides unit polytope is a \emph{triangle}, i.e.\ that $\Delta$ is
simplicial.  This is false.  The both-sides unit condition is self-dual: it holds
for $\Delta$ if and only if it holds for $\Delta^\circ$.  Were every
two-dimensional face of a both-sides unit polytope a triangle, applying this to
both $\Delta$ and $\Delta^\circ$ would make $\Delta$ simplicial and simple at once,
hence a simplex --- contradicting the existence of both-sides unit polytopes with
more than five vertices (there is one with nine, Example~\ref{ex:pair}).  The
non-triangular faces of a both-sides unit polytope are genuine; the content of
Conjecture~\ref{conj:main} is that they are \emph{smoothable}
(centrally symmetric), never of the new non-smoothable types.
\end{remark}

\begin{example}[the both-sides examples are mirror pairs]\label{ex:pair}
Among the $3{,}905{,}789$ reflexive $4$-polytopes with at most nine vertices,
exactly five are both-sides unit; they are closed under polar duality (two polar
pairs and one self-polar polytope).  One pair is the smooth simplex and the
simplex $\Delta = \conv\{(1,0,0,0),(0,1,0,0),(2,4,5,0),(3,3,0,5),(-6,-8,-5,-5)\}$,
whose ten two-dimensional faces are all rigid triangles (cyclic quotient points)
and whose polar is smooth.  A second pair consists of a simplicial six-vertex
polytope, all of whose eighteen faces are smooth triangles, and its polar (nine
vertices), whose fifteen faces are six rigid triangles and nine
centrally-symmetric quadrilaterals of type \textup{(S)} with two interior points.
In every case the two-dimensional faces are triangles or zonotopes, and every
non-smoothable member is paired under the mirror with a \emph{smooth} threefold,
not a singular one.  These are exactly the classical quotient examples one
dimension beyond which \cite{paper1} was written.
\end{example}

\begin{example}[a both-sides pair at the top of the classification]
\label{ex:pair36}
Both-sides unit polytopes are not confined to few vertices.  The $36$-vertex
product $H \times H$ of two reflexive hexagons --- the reflexive
$4$-polytope with the most vertices --- is both-sides unit, and its polar,
the $12$-vertex free sum $H^\circ \oplus H^\circ$, is the mirror partner
guaranteed by self-duality of the both-sides condition.  The faces of
$H \times H$ are $36$ unit parallelograms and $12$ centrally-symmetric
$\mathrm{dP}_6$ hexagons --- all zonotopes, so
Conjecture~\ref{conj:main} holds --- and the corresponding $X$ is singular
with $48$ isolated points, all of type \textup{(S)} ($36$ ordinary double
points and $12$ $\mathrm{dP}_6$-cone points).  The faces of the polar are
$72$ smooth triangles, so the mirror $X^\circ$ is a \emph{smooth}
Calabi--Yau threefold.  The pattern of Example~\ref{ex:pair} repeats at the
opposite end of the classification: a singular member (here everywhere
locally smoothable) mirrors a smooth one.  All assertions are machine-checked
in \texttt{missing\_polytope.py}.
\end{example}

\subsection{Computational evidence}
The exact-arithmetic classifier of \cite{paper1}, applied to a two-dimensional
face together with its two facet normals, computes $\ell(F^\circ)$ and the type of
$F$; a scan over the Kreuzer--Skarke list \cite{KreuzerSkarke00} then tests
Conjecture~\ref{conj:main} directly.  Over all $3{,}905{,}789$ reflexive
$4$-polytopes with at most nine vertices we find no counterexample: exactly
five polytopes are both-sides unit, and every
two-dimensional face of each is a triangle or a
centrally-symmetric zonotope.  (Both scanners --- the reference one and a
fast-engine one capable of the higher vertex counts --- are among the
ancillary files; every both-sides hit of the fast scanner is re-verified by
the reference toolkit.)

Three further pieces of data come from the full classification.  First, the
complete sweep of \cite{paper1} tracked, for every polytope with at least
$23$ vertices ($3{,}735{,}806$ of them), whether it is both-sides unit
\emph{and} carries a non-smoothable face: none is.  So a counterexample to
Conjecture~\ref{conj:main} with a \emph{non-smoothable} asymmetric face ---
the only kind relevant to Corollary~\ref{cor:mirror} --- has, in addition,
no instance with $v \ge 23$.  Second, the unique reflexive $4$-polytope
with more than $33$ vertices --- the $36$-vertex product $H \times H$ of two
reflexive hexagons, the one polytope absent from the per-vertex-count mirror
used in \cite{paper1} --- is both-sides unit, and conforms: its $48$
two-dimensional faces are $36$ unit parallelograms and $12$
$\mathrm{dP}_6$ hexagons, all centrally symmetric (ancillary file
\texttt{missing\_polytope.py}).  Third, its polar --- the $12$-vertex free
sum $H^\circ \oplus H^\circ$ --- is then also both-sides unit, with all $72$
two-dimensional faces smooth triangles; see
Example~\ref{ex:pair36}.

\subsection{Reduction to a local lemma}
The obstruction underlying Conjecture~\ref{conj:main} is local to a face and its
star.  Say a two-dimensional face $F$ is \emph{asymmetric} if it is neither a
triangle nor a zonotope, and call the two-dimensional faces sharing at least one
vertex with $F$ the \emph{star} of $F$.

\begin{conjecture}[local balancing]\label{conj:local}
Let $F$ be an asymmetric two-dimensional face of a reflexive $4$-polytope such
that $F$ and every face in its star have unit edges, and $\ell(F^\circ)=1$.  Then
some face in the star of $F$ has $\ell(\cdot^\circ) \ge 2$.
\end{conjecture}

Conjecture~\ref{conj:local} implies Conjecture~\ref{conj:main}: on a both-sides
unit polytope all edges are unit and all dual-edge lengths equal $1$, so no
asymmetric face can occur.  It is strictly local --- it constrains only $F$ and
its star, ignoring the rest of $\Delta$ --- and correspondingly it is a stronger
statement, with far more instances.  Over the reflexive $4$-polytopes with at most
eight vertices ($1{,}037{,}834$ polytopes) there are $38{,}571$ asymmetric faces
with a unit-edge star and $\ell(F^\circ)=1$, and Conjecture~\ref{conj:local} holds
for every one of them: each has a star face with $\ell(\cdot^\circ)\ge 2$.

Two structural facts point toward a proof.  First (Lemma~\ref{lem:mult}), for a
face $F$ with an interior lattice point $m$, every facet of $\Delta$ through an
edge $e$ of $F$ other than the two facets through $F$ has inner-normal
$(x_1,x_2)$-part a positive multiple of the outer edge-normal $e^\perp$: the
facets adjacent to $F$ across its edges lean strictly inward.  Second, comparing a
zonotope face admitting an all-$\ell{=}1$ star with an asymmetric face shows the
former places each edge of $F$ on the minimal number ($3$) of facets, while the
latter forces some edge onto more, carrying the excess into a dual edge of length
$\ge 2$.  A rigorous local proof --- in particular of the quadrilateral case,
where ``asymmetric'' means ``rigid'' --- is the natural next step and is left
open.

\section{Consequences and questions}\label{sec:questions}

Corollary~\ref{cor:mirror} follows from Proposition~\ref{prop:corr} and
Conjecture~\ref{conj:main}: a non-quotient non-smoothable germ $C(F)$ has $F$
non-triangular and of type \textup{(R)} or \textup{(D)}, hence not a zonotope, so
$\Delta$ is not both-sides unit and $X^\circ$ is non-isolated.  The two flagship
germs of \cite{paper1} are covered: the anticanonical cone over $\FF_1$ is the
rigid quadrilateral, and the deformable-but-non-smoothable cone is the pentagon
$Q = T + s$; neither is a triangle or a zonotope.

\begin{remark}
The unconditional part of the story is already definite.  By
Proposition~\ref{prop:corr} and Lemma~\ref{lem:mult}, whenever a unit-edge
$\Delta$ has \emph{any} two-dimensional face $F$ with $\ell(F^\circ) \ge 2$ the
mirror $X^\circ$ is non-isolated; and the both-sides scan shows this is the
generic situation --- among the unit-edge reflexive $4$-polytopes with at most
nine vertices that carry a non-smoothable face, all but two have $\max_F
\ell(F^\circ) \ge 2$, and the two exceptions are both-sides unit with only
triangular (quotient) non-smoothable faces; among the polytopes with at
least $23$ vertices the sweep of \cite{paper1} found no both-sides polytope
with a non-smoothable face at all.  Conjecture~\ref{conj:main} upgrades
these two exceptions to ``the exceptions are exactly the quotient stratum.''
\end{remark}

\begin{question}
Prove Conjecture~\ref{conj:local} (hence Conjecture~\ref{conj:main}).  The
quadrilateral case asks: can a rigid unit-edge quadrilateral occur as a
two-dimensional face of a reflexive $4$-polytope with $\ell(F^\circ)=1$ and a
unit, all-$\ell{=}1$ star?  The evidence says no.
\end{question}

\begin{question}
Corti--Filip--Petracci \cite{CFP} relate the smoothing components of $C(Q)$ to the
$0$-mutable Laurent polynomials with Newton polygon $Q$, a mirror-theoretic
statement, and Filip \cite{Filip25} treats the non-isolated case.  In that
language, what is the mirror-theoretic meaning of a face of type \textup{(R)} or
\textup{(D)}?  Corollary~\ref{cor:mirror} says these are precisely the faces whose
mirror is non-isolated; one expects non-smoothability to appear as the absence of
a $0$-mutable Laurent polynomial, and it would be interesting to read the
trichotomy off the canonical scattering diagram of the mirror.
\end{question}

\begin{thebibliography}{99}

\bibitem{Altmann97}
K.~Altmann,
\emph{The versal deformation of an isolated toric Gorenstein singularity},
Invent.\ Math.\ \textbf{128} (1997), no.~3, 443--479.

\bibitem{Batyrev94}
V.~V. Batyrev,
\emph{Dual polyhedra and mirror symmetry for Calabi--Yau hypersurfaces in toric
varieties},
J.\ Algebraic Geom.\ \textbf{3} (1994), 493--535.

\bibitem{CFP}
A.~Corti, M.~Filip, and A.~Petracci,
\emph{Mirror symmetry and smoothing Gorenstein toric affine 3-folds},
in \emph{Facets of Algebraic Geometry}, vol.~I, London Math.\ Soc.\ Lecture Note
Ser.\ \textbf{472}, Cambridge Univ.\ Press, 2022, pp.~132--163; arXiv:2006.16885.

\bibitem{Filip25}
M.~Filip,
\emph{Laurent polynomials and deformations of non-isolated Gorenstein toric
singularities},
arXiv:2504.04486 (2025).

\bibitem{KreuzerSkarke00}
M.~Kreuzer and H.~Skarke,
\emph{Complete classification of reflexive polyhedra in four dimensions},
Adv.\ Theor.\ Math.\ Phys.\ \textbf{4} (2000), 1209--1230.

\bibitem{paper1}
B.~J. Wuebben,
\emph{Non-smoothable Calabi--Yau threefolds from reflexive polytopes},
preprint, 2026.

\bibitem{paper2}
B.~J. Wuebben,
\emph{Smoothing Calabi--Yau threefolds in Gorenstein toric Fano fourfolds},
preprint, 2026.

\end{thebibliography}

\bigskip
\noindent{\sc Bernd J. Wuebben, New York, NY,} \texttt{wuebben@gmail.com}

\end{document}
